A Wait-free Queue with Polylogarithmic Step Complexity

### 2022-12-31 podc_paper/algorithm.tex
working on alg description, tidying code

\subsection{Overview of Implementation}
We use a \emph{tournament tree} to agree on a total ordering of the operations performed on the queue.
The tree is a static binary tree of height $\ceil{\log_2 p}$ with one leaf 
assigned to each process. 
Each tree node  stores an array of \emph{blocks}, where each block represents a 
sequence of enqueues and a sequence of dequeues.
See Figure \ref{tournament} for an example.
In this section, we assume for simplicity that each node has an infinite blocks array.
Section \ref{reducing} describes how to replace the infinite array by a representation that uses bounded space.

@@

Now suppose $P$'s operation has been propagated all the way to the root.
If $P$'s operation is an enqueue, it has obtained a place in the linearization ordering and can terminate.
If $P$'s operation is a dequeue, $P$ must use information in the tree to compute the value that the
dequeue must return.  To do this, $P$ first determines which block in the root contains its operation
(since the operation may have been propagated to the root by some other process).
Then, $P$ determines whether the queue is empty when its dequeue is linearized. 
If so, it returns \var{null} and we call it a \emph{null dequeue}.
If not, $P$ computes the rank\footnote{We say that the $r$th element in a sequence has rank $r$ within that sequence.} $r$ of its dequeue in the linearization ordering
among all non-null dequeues,
finds the $r$th enqueue in the linearization, and returns that enqueue's value.

@@

\begin{figure}[tb]
\input{tournamentTree.pdf_t}
\caption{An example tournament tree with four processes. 
We show explicitly the enqueue sequence and dequeue sequence represented by each block in the \var{blocks} arrays of the seven nodes.  The leftmost element of each \var{blocks} array is a dummy block.
Arrows represent the indices stored in \eleft\ and \eright\ fields of blocks (as described in Section \ref{sec:fields}).
The fourth process's Deq\sub{6} is still in progress.
The linearization order for this tree is
Enq(a) Enq(e) Deq\sub{2} $\mid$ Enq(b) Deq\sub{4} Deq\sub{5} $\mid$ Enq(d) Enq(f) Enq(h) Deq\sub{1} $\mid$ Enq(c) Deq\sub{3} $\mid$ Enq(g), where vertical bars indicate boundaries of blocks in the root.\label{tournament}}
\end{figure}

\begin{figure}
\input{implicit.pdf_t}
\caption{\label{implicit}The actual, implicit representation of the tree shown in Figure \ref{tournament}.
The leaf blocks simply show the \var{element} field.
Internal blocks show the \var{sum\sub{enq}} and \var{sum\sub{deq}} fields,
and \eleft\ and \eright\ fields are shown using arrows as before.
Root blocks also contain the additional \var{size} field.
The \var{super} field is not shown.}
\end{figure}

@@

Each \node\ of the tournament tree has an infinite array called \var{blocks}.
To simplify the code, each \node\ has an empty \block\ in \var{blocks[0]}.
Each \node\ also has a \var{head} index that stores the position in the \var{blocks} array to be used
for the next attempt to append a \block\ to this \node.

@@

The information stored in a \block\ depends on whether it is in an internal node or a leaf.
A block in a leaf represents a single enqueue or dequeue.  The \var{element} field stores the value
enqueued for an enqueue operation, or \var{null} if the operation is a dequeue.

@@

Each block in an internal \node\ $v$ stores the indices of the last direct subblock in $v$'s left and right child in the fields \eleft\ and \eright.  This allows us to navigate to the direct subblocks of any block easily.
Blocks also store prefix sums of the numbers of enqueues and dequeues:
the block in $v.\var{blocks}[i]$ has two fields \var{sum\sub{enq}} and \var{sum\sub{deq}}
that store the total number of enqueues and dequeues in $v.\var{blocks[1..i]}$.
These fields allow us to pinpoint the exact location of an operation among the subblocks of a given block.
For example, consider finding the $r$th enqueue in the linearization ordering.
We use the \var{sum\sub{enq}} fields of the root's blocks to do a binary search
to locate which block in the root contains the operation.
If we know which block $B$ in some \node\ $v$ contains the enqueue,
we can use the \var{sum\sub{enq}} field again to determine which child of $v$ contains the enqueue
and then to do a binary search
among the direct subblocks of $B$ in that child.
Thus, we work our way down the tree until we find a leaf block, which explicitly stores 
the enqueue.
%Now, suppose we want to find the $r$th enqueue in the linearization ordering for task \ref{findenqueue}.Let $B_1, B_2, \ldots, B_k$ be the blocks in the root.
%First, we need prefix sums of the number of enqueues in $E(B_1)\cdot E(B_2)\cdots E(B_i)$
%so that we can do a binary search for the block $B_e$ that contains the $r$th enqueue.
%This prefix sum also allows us to know the rank $r'$ within $E(B_e)$ of the $r$th enqueue.
%Once we have $r'$, we need the number of enqueues that $B_e$ received from its left child
%to determine whether the enqueue came from the left or right child of the root.
%Suppose the enqueue came from the right child $v_r$.
%Then, we know that the index of the block in $v_r$ that contains the enqueue
%is between $B_{e-1}.\eright + 1$ and $B_e$.\eright.
%We can again do a binary search within this range.
%For this, we can again use the prefix sums of the number of enqueues in any prefix of the array $v_r.blocks$.
%We can then continue in this way down the tree until reaching a leaf where the enqueue is stored explicitly.
We shall show that the binary search in the root can be done in $O(\log p + \log q)$ steps,
and the binary search within each other \node\ along the path to a leaf can be done in $O(\log p)$ steps,
for a total of $O(\log^2 p + \log q)$ steps to complete task \ref{findenqueue}.
All of the information needed for this search process can be derived from the 
\eleft, \eright\ and \var{sum\sub{enq}} fields.

To facilitate task \ref{findinroot}, each block has a field \var{super} that contains
the (approximate) index of its superblock in the parent \node's \var{blocks} array.
We will ensure that this field's value differs from the true index of the superblock by at most 1.
This allows a process to determine the true location of the superblock by checking the \eleft\ or \eright\ values of just two \blocks\ in the parent \node.
Thus, starting from an operation in a leaf's block, one can use these indices to track the 
operation all the way up the path to the root, and determine the operation's location in a root block
in $O(\log p)$ time.

We now consider task \ref{findrank}.
First, we must determine whether the queue is empty when a dequeue occurs.
To facilitate this, each block in the root stores a \var{size} field that stores the number of elements
in the queue after all operations in the linearization ordering up to that block (inclusive) 
have been performed.
We can easily determine which dequeues in a block $B_d$ in the root are null dequeues using
$B_{d-1}.\var{size}$, which stores the size of the queue just before $B_d$'s operations are performed, and the number of enqueues and dequeues in $B_d$.
We can also compute the number of non-null dequeues in blocks $B_1, \ldots, B_{d-1}$ 
as $B_{d-1}.\var{sum\sub{enq}}-B_{d-1}.\var{size}$.
Then, for any non-null dequeue in $B_d$, we can use this information to determine its
rank among all non-null dequeues in the linearization ordering, and hence the rank of the enqueue
whose value it should return (among all enqueues).

Now that we have defined the fields required for tasks \ref{findinroot}, \ref{findrank} and \ref{findenqueue},
we can easily see how to construct a new block $B$ during a \op{refresh} in $O(1)$ time.
A \op{refresh} on \node\ $v$ reads the value $h$ of the \var{head} field of each of $v$'s children and stores 
$h-1$ in the $B.\eleft$ and $B.\eright$ fields of the new block.
Then, we can compute $B.\var{sum\sub{enq}}$ as $v.\var{left}.\var{blocks}[B.\eleft].\var{sum\sub{enq}} + v.\var{right}.\var{blocks}[B.\eright].\var{sum\sub{enq}}$.
For a block $B$ in the root, $B.\var{size}$ can be computed using the \var{size} field of the previous block $B'$ and
the number of enqueues and dequeues in $B$:
$B.\var{size} = \max(0, B'.\var{size} + (B.\var{sum\sub{enq}}-B'.\var{sum\sub{enq}}) - (B.\var{sum\sub{deq}} - B'.\var{sum\sub{deq}}))$.

The only remaining field is $B.\var{super}$.  When the block 
$B$ is created for a \node\ $v$, we do not yet know where its
superblock will eventually be installed in $B$'s parent.
So, we leave this field blank.  After $B$ is installed 
in $v.\var{blocks[i]}$, processes cooperate to fill it in 
when they attempt to advance $v.\var{head}$ from $i$ to $i+1$.
They use the value read from the \var{head} field of $v$'s parent at that time.
As mentioned above, this might not be exactly the right index for $B$'s superblock, but we
shall prove that it is close.

\renewcommand{\algorithmiccomment}[1]{\hfill\eqparbox{COMMENTSINGLE}{\com\ #1}}
\begin{figure}
\here{Does \var{leaf} have to be shared for helping when doing GC?}
\begin{algorithmic}[1]
\setcounter{ALG@line}{1}
\Statex \linecomment{Shared variable}
\begin{itemize}
\item \var{Node} \var{root} \Comment{root of a binary tree of \tt{Node}s with one leaf for each process}
\end{itemize}

\Statex \linecomment{Thread-local variable}
\begin{itemize}
\item \var{Node} \var{leaf} \Comment{process's leaf in the tree}
\end{itemize}

\Statex $\blacktriangleright$ \var{Node}
\begin{itemize}
\item \var{Node left, right, parent} \Comment{children and parent pointers initialized  when creating the tree}
\item \var{Block[0..$\infty$] blocks} \Comment{initially \var{blocks[0]} contains an empty block with all integer fields equal to 0}
\item \var{int \head} \Comment{position to attempt appending next \block\ to \var{blocks}, initially 1}
\end{itemize}

\Statex $\blacktriangleright$ \var{Block}

\begin{itemize}
  	\item \var{int sum\sub{enq}, sum\sub{deq}}
  		\Comment{number of enqueue, dequeue operations in \var{blocks} array up to this block (inclusive)}
  	\item \var{int super}
  		\Comment{approximate index of the block's superblock in \var{parent.blocks}}
	\item[\com] Blocks in internal \nodes\ have the following additional fields
	\item \var{int} \eleft, \eright
  		\Comment{index of last direct subblock in the left and right child}
  	\item[\com] Blocks in leaf \nodes\ have the following additional field
  	\item \var{Object element}
  		\Comment{if the block's operation is \op{Enqueue}(\var{x}) then \var{element=x}, otherwise \var{element=null}.}
	\item[\com] Blocks in the root \node\ have the following additional field
	\item \var{int \size}%
  		\Comment{size of the queue after performing all operations up to the end of this block}
\end{itemize}

@@

We now discuss the queue implementation in more detail.  Pseudocode is provided in Figure \ref{pseudocode1} and \ref{pseudocode2}.
We use $v.\var{blocks[i].num\sub{enq}}$ as shorthand for 
$v.\var{blocks[i].sum\sub{enq} - blocks[i-1].sum\sub{enq}}$, that is, 
the number of enqueues in the block.  (For $\var{i}=0$, $v.\var{blocks[0].num\sub{enq}} = 0$.)
We use \var{num\sub{deq}} similarly.
\here{Check if the num abbreviation is really needed in the code--how many times do we use it?}

@@

\Function{void}{Enqueue}{\var{Object e}} 
    \State \hangbox{let \var{new} be a new \var{Block} with \var{element} \assign\ \var{e},\\
		\var{sum\sub{enq} \assign\ leaf.blocks[leaf.\head-1].sum\sub{enq}+1},\\
		\var{sum\sub{deq} \assign\ leaf.blocks[leaf.\head-1].sum\sub{deq}}}
    \State \Call{Append}{new}
\EndFunction{Enqueue}

@@

\Function{Object}{Dequeue()}{} 
    \State \hangbox{let \var{n} be a new \var{Block} with \var{element} \assign\ \var{null},\\
	    \var{sum\sub{enq} \assign\ leaf.blocks[leaf.\head-1].sum\sub{enq}},\\
	    \var{sum\sub{deq} \assign\ leaf.blocks[leaf.\head-1].sum\sub{deq}+1}}
    \State \Call{Append}{n}
    \State \var{<b, i>} \assign\ \Call{IndexDequeue}{leaf, leaf.\head, 1}
    \State \Return{ \Call{FindResponse}{b, i}}
    \label{deqRest}
\EndFunction{Dequeue}

@@

\Function{void}{Append}{\var{Block B}} 
    \State \linecomment{append block to leaf and propagate to root}
    \State \var{leaf.blocks[leaf.\head] \assign\ B} \label{appendLeaf}
    \State \var{leaf.\head\ \assign\ leaf.\head+1} \label{appendEnd} 
    \State \Call{Propagate}{\var{leaf.parent}} 
\EndFunction{Append}

@@

\Function{void}{Propagate}{\var{Node v}}
    \State \linecomment{propagate blocks from \var{v}'s children to root}
    \If{\bf{not} \Call{Refresh}{v}} \label{firstRefresh}  \hfill \com\ double refresh
        \State \Call{Refresh}{v} \label{secondRefresh}
    \EndIf
    \If{\var{v} \bf{is not} \var{root}} \hfill \com\ recurse up tree
        \State \Call{Propagate}{v.parent}
    \EndIf
\EndFunction{Propagate}

@@

\Function{boolean}{Refresh}{\var{Node v}}
    \State \linecomment{try to append a new block to \var{v.blocks}}
    \State \var{h \assign\ v.\head} \label{readHead}
    \ForEach{\var{dir} {\keywordfont{in}} \var{\{left, right\}}} \label{startHelpChild1}
        \State \var{h\sub{dir} \assign\ v.dir.\head} \label{readChildHead}
        \If{\var{v.dir.blocks[h\sub{dir}] $\neq$ null}} \label{ifHeadnotNull}
            \State{\Call{Advance}{v.dir, h\sub{dir}}} \label{helpAdvance}
        \EndIf
    \EndFor \label{endHelpChild1}
    \State \var{new} \assign\ \Call{CreateBlock}{v,h} \label{invokeCreateBlock}
    \If{\var{new = null}} \Return{\var{true}} \label{addOP} 
    \EndIf
    \State{\var{result} \assign\ \var{CAS(v.blocks[h], null, new)}} \label{cas}
    \State{\var{\Call{Advance}{v, h}}} \label{advance}
    \State \Return{ \tt{result}}
\EndFunction{Refresh}

@@

\Function{void}{Advance}{\var{Node v, int h}} 
    \State \linecomment{set \nf{v.blocks[h].super} and }
    \State \linecomment{increment \nf{v.head} from \nf{h} to \nf{h+1}}
    \State \var{h\sub{p}} \assign\ \var{v.parent.\head} \label{readParentHead}
    \State \var{CAS(v.blocks[h].super, null, h\sub{p})} \label{setSuper1}
    \State \var{CAS(v.head, h, h+1)} \label{incrementHead}
\EndFunction{Advance}

@@

\Function{Block}{CreateBlock}{\var{Node v, int i}} 
    \State\linecomment{create new block for a \op{Refresh} to install in \var{v.blocks[i]}}
    \State\linecomment{(return \var{null} if no blocks need to be propagated to \var{v})}
    \State let \var{new} be a new \var{Block} \label{initNewBlock}
    \State \var{new.\eleft} \assign\ $\var{v.left.head} - 1$
    \State \var{new.\eright} \assign\ $\var{v.right.head} - 1$
	\State \hangbox{\var{new.sum\sub{enq}} \assign\ \var{v.left.blocks[new.\eleft].sum\sub{enq}} + \\
			\var{v.right.blocks[new.\eright].sum\sub{enq}}}
	\State \hangbox{\var{new.sum\sub{deq}} \assign\ \var{v.left.blocks[new.\eleft].sum\sub{deq}} + \\
			\var{v.right.blocks[new.\eright].sum\sub{deq}}}
    \State \var{num\sub{enq}} \assign\ \var{new.sum\sub{enq} - v.blocks[i-1].sum\sub{enq}}
    \State \var{num\sub{deq}} \assign\ \var{new.sum\sub{deq} - v.blocks[i-1].sum\sub{deq}}
    \If{$\var{v} = \var{root}$}
        \State \hangbox{\var{new.size \assign\ max(0, v.blocks[i-1].\size\ + num\sub{enq} - num\sub{deq})}\label{computeLength}}
    \EndIf
    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}
        \State \Return \var{null}
    \Else
        \State \Return \var{new}
    \EndIf
\EndFunction{CreateBlock}

@@

\Function{element}{GetEnqueue}{\var{Node v, int b, int i}} \Comment{returns argument of ith enqueue in $E(\var{v.blocks[b]})$}
    \State \linecomment{Precondition: $\var{i}\geq 1$ and $E(\var{v.blocks[b]})$ contains at least \var{i} enqueues}

>>>>

\subsection{Overview}
We use a \emph{tournament tree} to agree on a total ordering of the operations performed on the queue.
The tree is a static binary tree of height $\ceil{\log_2 p}$ with one leaf 
assigned to each process. 
Each tree node  stores an array of \emph{blocks}, where each block represents a 
sequence of enqueues and a sequence of dequeues.
See Figure \ref{tournament} for an example.
In this section, we assume for simplicity that each node has an infinite blocks array.
Section \ref{reducing} describes how to replace the infinite array by a representation that uses bounded space.

@@

Now suppose $P$'s operation has been propagated all the way to the root.
If $P$'s operation is an enqueue, it has obtained a place in the linearization ordering and can terminate.
If $P$'s operation is a dequeue, $P$ must use information in the tree to compute the value that the
dequeue must return.  To do this, $P$ first determines which block in the root contains its operation
(since the operation may have been propagated to the root by some other process).
Then, $P$ determines whether the queue is empty when its dequeue is linearized. 
If so, it returns \nl\ and we call it a \emph{null dequeue}.
If not, $P$ computes the rank\footnote{We say that the $r$th element in a sequence has rank $r$ within that sequence.} $r$ of its dequeue in the linearization ordering
among all non-null dequeues,
finds the $r$th enqueue in the linearization, and returns that enqueue's value.

@@

\begin{figure}[tb]
\input{tournamentTree.pdf_t}
\caption{An example tournament tree with four processes. 
We show explicitly the enqueue sequence and dequeue sequence represented by each block in the \fld{blocks} arrays of the seven nodes.  The leftmost element of each \fld{blocks} array is a dummy block.
Arrows represent the indices stored in \eleft\ and \eright\ fields of blocks (as described in Section \ref{sec:fields}).
The fourth process's Deq\sub{6} is still in progress.
The linearization order for this tree is
Enq(a) Enq(e) Deq\sub{2} $\mid$ Enq(b) Deq\sub{4} Deq\sub{5} $\mid$ Enq(d) Enq(f) Enq(h) Deq\sub{1} $\mid$ Enq(c) Deq\sub{3} $\mid$ Enq(g), where vertical bars indicate boundaries of blocks in the root.\label{tournament}}
\end{figure}

\begin{figure}
\input{implicit.pdf_t}
\caption{\label{implicit}The actual, implicit representation of the tree shown in Figure \ref{tournament}.
The leaf blocks simply show the \fld{element} field.
Internal blocks show the \fld{sum\sub{enq}} and \fld{sum\sub{deq}} fields,
and \eleft\ and \eright\ fields are shown using arrows as before.
Root blocks also contain the additional \fld{size} field.
The \fld{super} field is not shown.}
\end{figure}

@@

Each \node\ of the tournament tree has an infinite array called \fld{blocks}.
To simplify the code, each \node\ has an empty \block\ in \fld{blocks[0]}.
Each \node\ also has a \fld{head} index that stores the position in the \fld{blocks} array to be used
for the next attempt to append a \block\ to this \node.

@@

The information stored in a \block\ depends on whether it is in an internal node or a leaf.
A block in a leaf represents a single enqueue or dequeue.  The \fld{element} field stores the value
enqueued for an enqueue operation, or \nl\ if the operation is a dequeue.

@@

Each block in an internal \node\ $v$ stores the indices of the last direct subblock in $v$'s left and right child in the fields \eleft\ and \eright.  This allows us to navigate to the direct subblocks of any block easily.
Blocks also store prefix sums of the numbers of enqueues and dequeues:
the block in $v.\fld{blocks}[i]$ has two fields \fld{sum\sub{enq}} and \fld{sum\sub{deq}}
that store the total number of enqueues and dequeues in $\var{v}.\fld{blocks}[1..\var{i}]$.
These fields allow us to pinpoint the exact location of an operation among the subblocks of a given block.
For example, consider finding the $r$th enqueue in the linearization ordering.
We use the \fld{sum\sub{enq}} fields of the root's blocks to do a binary search
to locate which block in the root contains the operation.
If we know which block $B$ in some \node\ $v$ contains the enqueue,
we can use the \fld{sum\sub{enq}} field again to determine which child of $v$ contains the enqueue
and then to do a binary search
among the direct subblocks of $B$ in that child.
Thus, we work our way down the tree until we find a leaf block, which explicitly stores 
the enqueue.
%Now, suppose we want to find the $r$th enqueue in the linearization ordering for task \ref{findenqueue}.Let $B_1, B_2, \ldots, B_k$ be the blocks in the root.
%First, we need prefix sums of the number of enqueues in $E(B_1)\cdot E(B_2)\cdots E(B_i)$
%so that we can do a binary search for the block $B_e$ that contains the $r$th enqueue.
%This prefix sum also allows us to know the rank $r'$ within $E(B_e)$ of the $r$th enqueue.
%Once we have $r'$, we need the number of enqueues that $B_e$ received from its left child
%to determine whether the enqueue came from the left or right child of the root.
%Suppose the enqueue came from the right child $v_r$.
%Then, we know that the index of the block in $v_r$ that contains the enqueue
%is between $B_{e-1}.\eright + 1$ and $B_e$.\eright.
%We can again do a binary search within this range.
%For this, we can again use the prefix sums of the number of enqueues in any prefix of the array $v_r.blocks$.
%We can then continue in this way down the tree until reaching a leaf where the enqueue is stored explicitly.
We shall show that the binary search in the root can be done in $O(\log p + \log q)$ steps,
and the binary search within each other \node\ along the path to a leaf can be done in $O(\log p)$ steps,
for a total of $O(\log^2 p + \log q)$ steps to complete task \ref{findenqueue}.
All of the information needed for this search process can be derived from the 
\eleft, \eright\ and \fld{sum\sub{enq}} fields.

To facilitate task \ref{findinroot}, each block has a field \fld{super} that contains
the (approximate) index of its superblock in the parent \node's \fld{blocks} array.
We will ensure that this field's value differs from the true index of the superblock by at most 1.
This allows a process to determine the true location of the superblock by checking the \eleft\ or \eright\ values of just two \blocks\ in the parent \node.
Thus, starting from an operation in a leaf's block, one can use these indices to track the 
operation all the way up the path to the root, and determine the operation's location in a root block
in $O(\log p)$ time.

We now consider task \ref{findrank}.
First, we must determine whether the queue is empty when a dequeue occurs.
To facilitate this, each block in the root stores a \fld{size} field that stores the number of elements
in the queue after all operations in the linearization ordering up to that block (inclusive) 
have been performed.
We can easily determine which dequeues in a block $B_d$ in the root are null dequeues using
$B_{d-1}.\fld{size}$, which stores the size of the queue just before $B_d$'s operations are performed, and the number of enqueues and dequeues in $B_d$.
We can also compute the number of non-null dequeues in blocks $B_1, \ldots, B_{d-1}$ 
as $B_{d-1}.\fld{sum\sub{enq}}-B_{d-1}.\fld{size}$.
Then, for any non-null dequeue in $B_d$, we can use this information to determine its
rank among all non-null dequeues in the linearization ordering, and hence the rank of the enqueue
whose value it should return (among all enqueues).

Now that we have defined the fields required for tasks \ref{findinroot}, \ref{findrank} and \ref{findenqueue},
we can easily see how to construct a new block $B$ during a \op{refresh} in $O(1)$ time.
A \op{refresh} on \node\ $v$ reads the value $h$ of the \fld{head} field of each of $v$'s children and stores 
$h-1$ in the $B.\eleft$ and $B.\eright$ fields of the new block.
Then, we can compute $B.\fld{sum\sub{enq}}$ as $v.\fld{left}.\fld{blocks}[B.\eleft].\fld{sum\sub{enq}} + v.\fld{right}.\fld{blocks}[B.\eright].\fld{sum\sub{enq}}$.
For a block $B$ in the root, $B.\fld{size}$ can be computed using the \fld{size} field of the previous block $B'$ and
the number of enqueues and dequeues in $B$:
$B.\fld{size} = \max(0, B'.\fld{size} + (B.\fld{sum\sub{enq}}-B'.\fld{sum\sub{enq}}) - (B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}}))$.

The only remaining field is $B.\fld{super}$.  When the block 
$B$ is created for a \node\ $v$, we do not yet know where its
superblock will eventually be installed in $B$'s parent.
So, we leave this field blank.  After $B$ is installed 
in $v.\fld{blocks}[\var{i}]$, processes cooperate to fill it in 
when they attempt to advance $v.\fld{head}$ from $i$ to $i+1$.
They use the value read from the \fld{head} field of $v$'s parent at that time.
As mentioned above, this might not be exactly the right index for $B$'s superblock, but we
shall prove that it is close.

\renewcommand{\algorithmiccomment}[1]{\hfill\eqparbox{COMMENTSINGLE}{\com\ #1}}
\begin{figure}
\here{Does \var{leaf} have to be shared for helping when doing GC?}
\begin{algorithmic}[1]
\setcounter{ALG@line}{1}
\Statex \linecomment{Shared variable}
\begin{itemize}
\item \typ{Node} \var{root} \Comment{root of a binary tree of \tt{Node}s with one leaf for each process}
\end{itemize}

\Statex \linecomment{Thread-local variable}
\begin{itemize}
\item \typ{Node} \var{leaf} \Comment{process's leaf in the tree}
\end{itemize}

\Statex $\blacktriangleright$ \typ{Node}
\begin{itemize}
\item \typ{Node} \fld{left}, \fld{right}, \fld{parent} \Comment{children and parent pointers initialized  when creating the tree}
\item \typ{Block}[0..$\infty$] \fld{blocks} \Comment{initially \var{blocks}[0] contains an empty block with all integer fields equal to 0}
\item \typ{int} \head \Comment{position to attempt appending next \block\ to \fld{blocks}, initially 1}
\end{itemize}

\Statex $\blacktriangleright$ \typ{Block}

\begin{itemize}
  	\item \typ{int} \fld{sum\sub{enq}, sum\sub{deq}}
  		\Comment{number of enqueue, dequeue operations in \fld{blocks} array up to this block (inclusive)}
  	\item \typ{int} \fld{super}
  		\Comment{approximate index of the block's superblock in \fld{parent.blocks}}
	\item[\com] Blocks in internal \nodes\ have the following additional fields
	\item \typ{int} \eleft, \eright
  		\Comment{index of last direct subblock in the left and right child}
  	\item[\com] Blocks in leaf \nodes\ have the following additional field
  	\item \typ{Object} \fld{element}
  		\Comment{if the block's operation is \opa{Enqueue}{x} then \fld{element}=\var{x}, otherwise \fld{element}=\nl}
	\item[\com] Blocks in the root \node\ have the following additional field
	\item \typ{int} \size%
  		\Comment{size of the queue after performing all operations up to the end of this block}
\end{itemize}

@@

We now discuss the queue implementation in more detail.  Pseudocode is provided in Figure \ref{pseudocode1} and \ref{pseudocode2}.
%We use $v.\var{blocks[i].num\sub{enq}}$ as shorthand for 
%$v.\var{blocks[i].sum\sub{enq} - blocks[i-1].sum\sub{enq}}$, that is, 
%the number of enqueues in the block.  (For $\var{i}=0$, $v.\var{blocks[0].num\sub{enq}} = 0$.)
%We use \var{num\sub{deq}} similarly.
%\here{Check if the num abbreviation is really needed in the code--how many times do we use it?}

@@

\Function{void}{Enqueue}{\typ{Object} \var{e}} 
    \State \hangbox{let \var{n} be a new \typ{Block} with \fld{element} \assign\ \var{e},\\
		\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{enq}}+1,\\
		\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{deq}}}
    \State \Call{Append}{\var{n}}
\EndFunction{Enqueue}

@@

\Function{Object}{Dequeue()}{} 
    \State \hangbox{let \var{n} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
	    \fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{enq}},\\
	    \fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{deq}}+1}
    \State \Call{Append}{\var{n}}
    \State \var{<b, i>} \assign\ \Call{IndexDequeue}{\var{leaf}, \var{leaf}.\head, 1}
    \State \Return{ \Call{FindResponse}{\var{b, i}}}
    \label{deqRest}
\EndFunction{Dequeue}

@@

\Function{void}{Append}{\typ{Block} \var{B}} 
    \State \linecomment{append block to leaf and propagate to root}
    \State \var{leaf}.\fld{blocks}[\var{leaf}.\head] \assign\ \var{B}\label{appendLeaf}
    \State \var{leaf}.\head\ \assign\ \var{leaf}.\head+1 \label{appendEnd} 
    \State \Call{Propagate}{\var{leaf}.\fld{parent}} 
\EndFunction{Append}

@@

\Function{void}{Propagate}{\typ{Node} \var{v}}
    \State \linecomment{propagate blocks from \var{v}'s children to root}
    \If{\bf{not} \Call{Refresh}{\var{v}}} \label{firstRefresh}  \hfill \com\ double refresh
        \State \Call{Refresh}{\var{v}} \label{secondRefresh}
    \EndIf
    \If{\var{v} \bf{is not} \var{root}} \hfill \com\ recurse up tree
        \State \Call{Propagate}{\var{v}.\fld{parent}}
    \EndIf
\EndFunction{Propagate}

@@

\Function{boolean}{Refresh}{\typ{Node} \var{v}}
    \State \linecomment{try to append a new block to \var{v}.\fld{blocks}}
    \State \var{h} \assign\ \var{v}.\head \label{readHead}
    \ForEach{\fld{dir} {\keywordfont{in}} \fld{\{left, right\}}} \label{startHelpChild1}
        \State \var{childHead} \assign\ \var{v}.\fld{dir}.\head \label{readChildHead}
        \If{\var{v}.\fld{dir.blocks}[\var{childHead}] $\neq$ \nl} \label{ifHeadnotNull}
            \State \Call{Advance}{\var{v}.\fld{dir}, \var{childHead}} \label{helpAdvance}
        \EndIf
    \EndFor \label{endHelpChild1}
    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlock}
    \If{\var{new = \nl}} \Return{\tr} \label{addOP} 
    \EndIf
    \State \var{result} \assign\ \Call{CAS}{\var{v}.\fld{blocks}[\var{h}], \nl, \var{new}} \label{cas}
    \State \var{\Call{Advance}{\var{v, h}}}\label{advance}
    \State \Return{ \var{result}}
\EndFunction{Refresh}

@@

\Function{void}{Advance}{\typ{Node} \var{v}, \typ{int} \var{h}} 
    \State \linecomment{set \var{v}.\fld{blocks}[\var{h}].\fld{super} and }
    \State \linecomment{increment \var{v}.\fld{head} from \var{h} to \var{h}+1}
    \State \var{h\sub{p}} \assign\ \var{v}.\fld{parent}.\head \label{readParentHead}
    \State \Call{CAS}{\var{v}.\fld{blocks}[\var{h}].\fld{super}, \nl, \var{h\sub{p}}} \label{setSuper1}
    \State \Call{CAS}{\var{v}.\head, \var{h}, \var{h}+1} \label{incrementHead}
\EndFunction{Advance}

@@

\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} 
    \State\linecomment{create new block for a \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]}
    \State\linecomment{(return \nl\ if no blocks need to be propagated to \var{v})}
    \State let \var{new} be a new \typ{Block} \label{initNewBlock}
    \State \var{new}.\eleft \assign\ $\var{v}.\fld{left}.\head - 1$
    \State \var{new}.\eright \assign\ $\var{v}.\fld{right}.\head - 1$
	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}}
	\State \hangbox{\var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}
    \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnq}
    \State \var{num\sub{deq}} \assign\ $\var{new}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
    \If{$\var{v} = \var{root}$}
        \State \hangbox{\var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \var{num\sub{enq}} - \var{num\sub{deq}}$)}\label{computeLength}
    \EndIf
    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}
        \State \Return \nl
    \Else
        \State \Return \var{new}
    \EndIf
\EndFunction{CreateBlock}

@@

\Function{element}{GetEnqueue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} \Comment{returns argument of \var{i}th enqueue in $E(\var{v}.\fld{blocks}[\var{b}])$}
    \State \linecomment{Precondition: $\var{i}\geq 1$ and $E(\var{v}.\fld{blocks}[\var{b}])$ contains at least \var{i} enqueues}
